Measuring Interaction Design

### 2015-09-19 paper.tex
introduction should be ok

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

\title{Measuring Interaction Design\\
Measuring Interaction before developing prototypes\\
Measuring Interaction Graphs}

@@

\begin{abstract}
  Early development of prototypes of user interfaces is good
  practice. However, they have to cover specific usage scenarios, and
  because they are limited in focus, the whole picture of the user
  interface is easily lost. Even simple questions dealing with number
  and length of interaction paths or impact of possible user errors
  can be answered only for the specific scenarios being analysed.

  We present a tool that transforms models of a user interface into a
  graph. This is then used to specify usage scenarios, and to generate
  possible interaction paths. Metrics based on possible paths, with or
  without possible mistakes, can be easily computed. When applying
  these metrics get precise and objective values that can be used to
  compare designs or to identify weakness. When applying them to mail
  applications, we learn for example that Gmail has \hl{7 times more
  optimal paths, it has 4x more paths include at most one possible
  user error, that it has 20\% fewer steps.}

@@

\begin{quote}
  Subcommittee: Technology, Systems, and Engineering

@@

Keyword we need to pay attention to:
This subcommittee will focus on technology, systems and engineering
contributions that enable, improve, or advance interaction. This will
include software and hardware technologies and systems that enable and
demonstrate novel interactive capabilities, as well as languages,
methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
systems. Engineering contributions should clearly demonstrate how they
address interactive systems concerns such as, for example,
scalability, reliability, interoperability, testing, and
performance. Systems and technology contributions will be judged by
their technical innovation and/or ability to connect, SIMPLIFY or
enrich interactions, for example in intelligent interfaces and
mobile/ubiquitous computing.

@@

We present an approach that allows a designer to assess interaction
design (IxD) qualities such as efficiency, consistency,
error-proneness and recovery from premature committment. Key
importance is given to the ability to quickly
\begin{inparaenum}[(a)]
\item understand how supportive a user interface (UI) is with respect
  to user efficiency;
\item understand how prone the UI is to user navigation
  errors;
\item understand how recoverable the UI is from those errors; and,
\item support task analysis and task/scenario design.
\end{inparaenum}
%
This allows different designs to be objectively compared to support
designers in the construction and engineering of interactive
systems. To explain our work we present a comparison of four different
web mail front ends. We have chosen this domain because it is very
well known and yet several trivial questions lead to non trivial
results. We applied the same techniques also in other domains, such as
HVAC and other embedded UIs.

The approach is based on UML statechart models of UIs which are
automatically processed to produce \emph{interaction graphs}. These
are then used to specify interaction scenarios and unfold interaction
paths (called \emph{execution traces}) determined by the specific
scenarios being considered. Traces are fed to several graph-theoretic
computations which produce a dashboard with different results with the
answers. Except for development of models and specification of the
desired scenarios, which have to be done manually, the other steps are
totally automatic; models of a design (such as the ones shown in the
paper) could be developed in a matter of a couple of hours.

Developing good UIs for web or mobile applications is a
complex and expensive endeavour. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centred Development practices is effective, as is
following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early
prototyping~\cite{buxton07} to explore part of the five-dimensional
prototyping space~\cite{mccurdy06}. It is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{rubin08,preece02}.

However, it requires prototypes which are usually developed with
certain tasks in mind, and therefore are quite restricted in terms of
depth, breadth, dynamics and data. Furthermore, in addition to the
possible bias introduced by prototypes, usability results are always
surrounded by a cloud of uncertainty, due to subjectivity introduced
by participants and facilitators or by other contingency factors
involved in the analysis. Thus, although a  significant effort
needs to be expended to develop and use prototypes, less than optimal
results are obtained.

Even worse, there might be questions whose answer cannot be easily
found. For example, given one or more potential IxD solutions and some
usage scenarios, interesting questions could include: ``How many
different ways can be followed by the user to carry out the scenario?'', ``Which
are the shortest ones?'', ``If a user makes a mistake, would he or she
be able to recover?'', ``How many steps would the recovery
require?''. When designing and evaluating embedded UIs (such as when
dealing with plane's cockpits~\cite{conversy07}), other relevant
questions might include ``How would the above properties change if we
add a certain a widget?'', or ``... if we replace a widget with
another?''.  At the moment, even these straightforward questions are
quite complicated to answer. In fact, they require development of
prototypes, inspecting them, manual tracking of which screens and
widgets are used at which stage, and a systematic manual analysis.

This should not be the case, however. These answers provide important
insights to a designer, and potentially support assements of potential user
flexibility, user efficiency, error proneness, ability to recover,
compactness and consistency of a design.

Our contribution consists of the development of a tool that transforms
models and scenario specifications into interaction paths, and the
definition of metrics that provide concise, precise and objective
measures of a design. The examples included in the paper show that
among four web mail applications, and with respect to a typical usage
scenario, Gmail is the most efficient and flexible UI, with the best
ability to recover from certain mistakes. In fact, Gmail has the
largest number of shortest sequences of steps (even when users are
supposed to make 1 or 2 mistakes), but when users make more than 2
mistakes the number of paths drops significantly; on average, best and worst
cases, Gmail features also shorter paths, requiring \hl{CHECK 20\% fewer steps};
however, the probability that a user hits an optimal path with Gmail
is almost half of that of another application, and the probability
that a random walker hits a state that is not due to a mistake is
\hl{SAY SOMETHING}.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
messages. If Gmail didn't exist yet, its designers could obtain these
answers well before performing usability studies.

@@

Gmail has the highest number of order 0 and 1 paths, and the highest
difference between the number of order 0 and 1, and between order 1
and 2 (at least 3 times as many order 0 paths than any of the other
applications, and at least twice as many order 1 paths as the any of
the other ones).  Gmail offers 3 times as many error-free alternative
paths to accomplish the scenario, which indicates that users might
more easily follow one of those paths. However, Gmail offers also
almost twice as many order 1 paths, which means that users could be
more easily induced into an erroneous path than when using another
system. Because the number of order 2 or 3 paths decreases, Gmail
reduces therefore the ``error proneness'' of this UI, for executions
that involve 2 or 3 errors. Notice that Roundcube has the smallest
number of order 0 paths (2 of them), which means that users are not
given much flexibility and freedom in carrying out correctly the
scenario.
For none of the systems a detour leads to dead-ends preventing the
user to complete the scenario.

Figure~\ref{fig:bestcase} shows the length of paths in the best case,
i.e., when users would always choose the shortest route. Gmail offers
the shortest paths across the four detour orders (for order 0 the
length is 13 steps; for order 3 the length is 17 steps; the other
applications are remarkably similar among them).  A plausible
interpretation is therefore that Gmail not only offers many more
error-free paths, but also gives the shortest ones. Users are given
more flexibility and more efficiency.

@@

To combining these two results, we can easily compute the frequency of
paths with different length. Figure~\ref{fig:freqopt} shows, for each
application, the frequency of order 0 paths, the frequency of paths
with length less or equal to 15 (the minimum length across the four
applications), and the frequency of optimal paths (the shortest
ones). These values show that Gmail users have the lowest probability
to hit an order 0 path (because of the relative large number of order
1 paths made available by Gmail), have the highest probability to hit
a path with length 15 or less, have the lowest probability of hitting
the shortest paths.  Thus, flexibility and efficiency that can be
exploited with Gmail is counterbalanced by the required knowledge and
capability of chosing an optimal path. In other words, Gmail offers
many detours of order 1 which increase flexibility for some users and
might decrease effectiveness for other ones.

% >   d=0.050 # 20% chances that a user makes a generic mistake
% >   dprh=detour.page.rank(sc.hom,3,d);dprh
%          0          1          2          3 
% 0.73455373 0.11735984 0.05684419 0.09124223 
% >   dprg=detour.page.rank(sc.gom,3,d);dprg
%          0          1          2          3 
% 0.69210344 0.17697177 0.06228221 0.06864257 
% >   dprs=detour.page.rank(sc.sqm,3,d);dprs
%          0          1          2          3 
% 0.79062415 0.07122957 0.05364190 0.08450438 
% >   dprr=detour.page.rank(sc.rcm,3,d);dprr
%          0          1          2          3 
% 0.76538478 0.08355996 0.06247253 0.08858273 %

Another probabilistic analysis can be performed using page rank, which
can be used to compute the probability that a random walk in a graph
visits a certain vertex. We computed the page rank (with a damping
value of 5\% - meaning that the randowm walker with probability 5\%
jumps to an arbitrary state and probability 95\% chooses one of the
actions available in the current state) for each vertex in the graph,
and then added the page rank value for vertices with different detour order.
Figure~\ref{fig:pr} shows the resulting probabilities.

@@

With Gmail the probability of visiting an order 0 state is close to
70\%, the lowest among the four applications. But when it comes to
visiting an order 0 or 1 state, the probability increases to 87\%,
which is the highest. Thus, to compare Gmail with SquirrelMail, a
completely random usage of SquirrelMail has 10\% more probability of
hitting an error-free state than Gmail. That advantage is reduced when
considering order 0 and 1 paths, because with SquirrelMail the
probability is 85\% and Gmail it is 87\%.

@@

@@

This means that users will have a twice as much large choice of
correct paths when composing a new message rather than ``simply''
replying to a read one. On average, when composing, users could be
9\% more efficient, and they are 4 times more likely to do the right
things. Thus, Horde is more suitable for composing new messages than
it is for replying to existing ones.

@@

Thus, one possible design question is ``What are the effects of adding
these functionalities?'' in a typical driving scenario where a speed
is selected, then the system is disengaged, and then the same speed
needs to be set?

System S (where we assumed that re-setting the speed is done manually
by the driver with 4 actions on ``up'' and ``down'' to get
approximately get the same speed) leads to 1 order 0 path requiring 7
steps; there are 8 order 1 paths with average length  9.4. The
probability of hitting the optimal path is 0.11; action density is
1.75, and unique action density is 1.25.

@@

Therefore we can conclude that:
\begin{inparaenum}
\item system A makes users more efficient;
\item with A there are two possible ways to achieve the scenario, thus
  more flexibility is given;
\item with A the probability of doing the right thing is almost half
  of system S: it might be more difficult to do the right thing
  because more possibilities are offered;
\item system A features a more compact design, with fewer action types
  to be performed at each moment, being therefore potentially easier
  to learn and remember.
\end{inparaenum}

\hl{present them in terms of :
consistency,
error proneness, efficiency, premature committment, flexibility}

\clearpage

@@

@@

@@

>>>>

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

\title{Measuring Interaction Design}

@@

\begin{abstract}
  Early prototyping of user interfaces is an established good
  practice. However, because prototypes cover only some usage
  scenarios, the whole picture of the user interface is easily
  lost.  Questions dealing with number of required steps,
  possible interaction paths or impact of possible user errors can be
  answered only for the specific scenarios being analysed and after
  tedious manual inspection.

  We present a tool that transforms models of a user interface into a
  graph, upon which usage scenarios can be easily specified, and used
  to compute possible interaction paths. Metrics based on possible
  paths, with or without possible user errors, can be easily computed.
  For example, when applying them to mail applications, we learn that
  Gmail has 3 times more shortest routes, has 2x more routes that
  include one user error, has routes with 13\% fewer steps, but has
  the smallest probability to hit an optimal route.

@@

\begin{quote}
\mytodo{ Subcommittee: Technology, Systems, and Engineering

@@

Keyword we need to pay attention to:
This subcommittee will focus on technology, systems and engineering
contributions that enable, improve, or advance interaction. This will
include software and hardware technologies and systems that enable and
demonstrate novel interactive capabilities, as well as languages,
methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
systems. Engineering contributions should clearly demonstrate how they
address interactive systems concerns such as, for example,
scalability, reliability, interoperability, testing, and
performance. Systems and technology contributions will be judged by
their technical innovation and/or ability to connect, SIMPLIFY or
enrich interactions, for example in intelligent interfaces and
mobile/ubiquitous computing.
}

@@

We present an approach that allows a designer to assess interaction
design (IxD) qualities such as efficiency,
error-proneness and recovery from premature committment. Key
importance is given to the ability to quickly
\begin{inparaenum}[(a)]
\item understand how supportive a user interface (UI) is with respect
  to user efficiency;
\item understand how prone the UI is to user navigation
  errors;
\item understand how recoverable the UI is from those errors; and,
\item perform objective quantitative comparison of different designs
  to support construction and engineering of interactive systems.
\end{inparaenum}

To explain our work we present a comparison of four web mail front
ends. We have chosen this domain because it is easy to understand and
yet several trivial questions are not easily answered. We applied
the same techniques also in other domains, such as HVAC and other
embedded UIs.

Developing good UIs for web or mobile applications is a
complex and expensive endeavour. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centered Development (UCD) practices is effective, as is
following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early
prototyping to explore part of the five-dimensional fidelity
prototyping space~\cite{mccurdy06,buxton07}. It is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{preece02,rubin08}.

However, UCD requires prototypes which are usually developed with
certain tasks in mind, and therefore are quite restricted in terms of
depth, breadth, dynamics and data. Furthermore, in addition to the
possible bias introduced by prototypes, usability results are always
surrounded by a cloud of uncertainty, due to subjectivity introduced
by participants and facilitators or by other contingency factors
involved in the analysis. Thus, although a  significant effort
needs to be directed to develop and use prototypes, less than optimal
results are obtained.

Even worse, there might be questions that cannot be cannot be easily
answered. Given one or more potential designs and some usage
scenarios, interesting questions could include: ``How many different
routes can be followed by the user to carry out the scenario?'',
``Which are the shortest ones?'', ``If a user makes a mistake, would
he or she be able to recover?'', ``How many steps would the recovery
require?''. In a different setting, when designing and evaluating
embedded UIs (such as when dealing with plane's
cockpits~\cite{conversy07}), other relevant questions might include
``How would the above properties change if we add a certain a
widget?'', or ``... if we replace a widget with another?''.  At the
moment, these straightforward questions are quite complicated to
answer. In fact, they require development of prototypes, inspecting
them, manual tracking of which screens and widgets are used at which
stage, and a systematic manual analysis.

This should not be the case, however. Answers to these questions could
be automatically computed. In this way a system could provide
important insights to a designer, and  support assements of
potential user flexibility, user efficiency, error proneness, ability
to recover, compactness and consistency of a design.

Our approach is based on UML statechart models of UIs which are
automatically processed to produce \emph{interaction graphs}. These
graphs are then used to specify interaction scenarios and to unfold
the possible interaction paths (called \emph{execution traces})
determined by the specific scenarios being considered. Traces are fed
to several graph-theoretic computations which produce a dashboard with
different results. Except for development of models and specification
of the desired scenarios, which have to be done manually, the other
steps are totally automatic; models of a design (such as the ones
shown below) can be developed in a matter of a couple of hours.

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a model to specify interaction scenarios,
\item the development of a tool (MIG-TOOL, Measuring Interaction
  Graphs Tool) that transforms models and scenario specifications into
  interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of a design.
\end{inparaenum}
The examples included in the paper show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, reducing the
error-proneness of the UI; on the best case, Gmail features also the
shortest paths, requiring 13\% fewer steps; however, the probability
that a user hits an optimal path with Gmail is 10 times smaller than
the best of the other applications, and the probability that a random
walker hits a state that is not due to an error is 10\% smaller than
the best of the other applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
messages. If Gmail didn't exist yet, by using MIGTOOL its designers
could obtain these answers well before developing prototypes and
performing usability studies.

@@

Gmail has the highest number of order 0 and 1 paths, and the highest
difference between the number of order 0 and 1, and between order 1
and 2 (at least 3 times as many order 0 paths than any of the other
applications, and at least twice as many order 1 paths as the any of
the other ones). 
 Gmail offers 3 times as many error-free alternative
paths to accomplish the scenario, which indicates that users might
more easily follow one of those paths. 
\mytodo{result}
However, Gmail offers also
almost twice as many order 1 paths, which means that users could be
more easily induced into an erroneous path than when using another
system.  
\mytodo{result}
Because the number of order 2 or 3 paths decreases, Gmail
reduces therefore the ``error proneness'' of this UI, for executions
that involve 2 or 3 errors. Notice that Roundcube has the smallest
number of order 0 paths (2 of them), which means that users are not
given much flexibility and freedom in carrying out correctly the
scenario. 
\mytodo{result}
For none of the systems a detour leads to dead-ends preventing the
user to complete the scenario.

Figure~\ref{fig:bestcase} shows the length of paths in the best case,
i.e., when users would always choose the shortest route. Gmail offers
the shortest paths across the four detour orders (for order 0 the
length is 13 steps, saving more than 13\% steps compared to other
applications; for order 3 the length is 17 steps; the other
applications are remarkably similar among them).  A plausible
interpretation is therefore that Gmail not only offers many more
error-free paths, but also gives the shortest ones. Users are given
more flexibility and more efficiency.  \mytodo{result} Because also
paths with order 1 or more are the shortest ones among the four
applications, Gmail also makes users more efficient in recovering from
errors.  \mytodo{result}

@@

To combine these two results, we can easily compute the frequency of
paths with different length. Figure~\ref{fig:freqopt} shows, for each
application, the frequency of order 0 paths, the frequency of paths
with length less or equal to 15 (the minimum length across the four
applications), and the frequency of optimal paths (the shortest
ones). These values show that Gmail users have the lowest probability
to hit an order 0 path (because of the relative large number of order
1 paths made available by Gmail), have the highest probability to hit
a path with length 15 or less, have the lowest probability of hitting
the shortest paths (10 times smaller than the best of the other
applications).  Thus, the flexibility and efficiency that can be
exploited with Gmail are counterbalanced by the required knowledge and
capability of chosing an optimal path. In particular, Gmail offers
many detours of order 1 which increase flexibility for some users and
might decrease effectiveness for less knowledgeable ones.
\mytodo{result}

% >   d=0.050 # 5% chances that a user makes a generic mistake
% >   dprh=detour.page.rank(sc.hom,3,d);dprh
%          0          1          2          3 
% 0.73455373 0.11735984 0.05684419 0.09124223 
% >   dprg=detour.page.rank(sc.gom,3,d);dprg
%          0          1          2          3 
% 0.69210344 0.17697177 0.06228221 0.06864257 
% >   dprs=detour.page.rank(sc.sqm,3,d);dprs
%          0          1          2          3 
% 0.79062415 0.07122957 0.05364190 0.08450438 
% >   dprr=detour.page.rank(sc.rcm,3,d);dprr
%          0          1          2          3 
% 0.76538478 0.08355996 0.06247253 0.08858273 %

Another probabilistic analysis can be performed using page rank, which
can be used to compute the probability that a random walk in a graph
visits a certain vertex~\cite{page1999pagerank}. We computed the page rank (with a damping
value of 5\% - meaning that the random walker with probability 5\%
jumps to an arbitrary state and probability 95\% chooses one of the
actions available in the current state) for each vertex in the graph,
and then added the page rank value for vertices with different detour order.
Figure~\ref{fig:pr} shows the resulting probabilities.

@@

With Gmail the probability of visiting an order 0 state is close to
70\%, the lowest among the four applications. But when it comes to
visiting an order 0 or 1 state, the probability increases to 87\%,
which is the highest. Thus, to compare Gmail with SquirrelMail, a
completely random usage of SquirrelMail has 10\% more probability of
hitting an error-free state than Gmail. That advantage is reduced when
considering order 0 and 1 paths, because with SquirrelMail the
probability is 85\% and Gmail it is 87\%.  This means that somebody
with no knowledge on how to use an email front end, when using Gmail
would have 10\% fewer chances of carrying out the scenario without
making any error, as opposed to when using SquirrelMail: SquirrelMail
provides more guidance. Across the four UIs, the probability of making
at most 1 error is approximately the same\footnote{These results are
  not very sensitive to the value chosen for the damping factor:
  differences across the four applications are very similar when $d$
  ranges in $[0.05,0.20]$.}.

\mytodo{result}

@@

\mytodo{result}

@@

This means that users will have a twice as much large choice of
correct paths when composing a new message rather than ``simply''
replying to a read one. On average, when composing, users could be
9\% more efficient, and they are 4 times more likely to do the right
things. Thus, Horde is more suitable for composing new messages than
it is for replying to existing ones.
\todo[inline]{result}

@@

Thus, one possible design question is ``What are the effects of adding
these functionalities?'' in a typical driving scenario where a speed
is selected, then the system is disengaged, and then the same speed
needs to be set.

System S (where we assumed that re-setting the speed is done manually
by the driver with 4 actions on ``up'' and ``down'' to
approximately get the same speed) leads to 1 order 0 path requiring 7
steps; there are 8 order 1 paths with average length  9.4. The
probability of hitting the optimal path is 0.11; action density is
1.75, and unique action density is 1.25.

@@

Therefore we can conclude that:
\begin{inparaenum}
\item system A makes users more efficient;
\item with A there are two possible ways to achieve the scenario, thus
  more flexibility is given;
\item with A the probability of doing the right thing is almost half
  of system S: it might be more difficult to do the right thing
  because more possibilities are offered;
\item system A features a more compact design, with fewer action types
  to be performed at each moment, being therefore potentially easier
  to learn and remember.
\end{inparaenum}
\todo[inline]{result}

@@

feasibility:

sensitivity with respect to models

@@

valdiity of results:

@@

MIGTOOL is part of a suite of model-based tools for designing,
analysing and testing user interfaces developed by the company
ANONYMIZED. Development of a web-based user interface of MIGTOOL is
underway; the company plans to make it freely available for research
purposes.
